#Exploratory Data Analysis for the Red Wine Quality Data Set by Joshua Maine
##10/6/18
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Install all the packages I use for the project. 
#I ran into repeated issues with packages and libraries if loaded previously. 
#Knitr constantly failed until I commented these out

#nstall.packages("ggplot2", dependencies = T, repos = "http://cran.us.r-project.org") 
#install.packages("knitr", dependencies = T, repos = "http://cran.us.r-project.org")
#install.packages("dplyr", dependencies = T, repos = "http://cran.us.r-project.org")
#install.packages("xtable", dependencies = T, repos = "http://cran.us.r-project.org")
#install.packages("ggally", dependencies =T, repos = "http://cran.us.r-project.org")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Calling all the libraries I might need
#library(ggplot2)
#library(dplyr)
#library(knitr)
#library(xtable)
#library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load the Data
wd <- read.csv('wineQualityReds.csv')
```


#INTRODUCTION


This dataset interest me as its similar to some further data analysis I want to 
do on data on beer. My longest running hobby, for almost a decade, has been 
homebrewing, and there is more similarities to red wines and beer than whites.


#Univariate Explorations and Plots


To get an idea of the data and gain a better understanding, I'm going to work 
through a few Univariate explorations. I'll start by getting an idea of the 
table structure and looking at the summary.


```{r echo=FALSE, message=FALSE, warning=FALSE}
str(wd)
```


The data set as 1599 records with 13 variables. Taking a full look at the 
variables, it has:

X appears to be the wine tested, the removal of name and other identifying 
information, shows this is likely from a judging or evaluation of the wines 
in the data set.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd)

```

With all the variables, this is most certainly data used in comparing and 
evaluting different red wines. Looking at the summary data there are
things I want to plot and get a better idea of how the data is grouped and to 
find any outliers for correction.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
qplot(x = quality, data = wd)
```

Most of the quality values exist as a 5 or a 6, so I may group these a little 
tigther to sort the standouts on both sides. Grouping the values by pairs to 
have below average, average, above average is shown below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Creating and ordering the rankings values
wd$ranking <- ifelse(wd$quality < 5, 'below average',
                     ifelse(wd$quality < 7, 'average', 'above average'))
wd$ranking <- ordered(wd$ranking,
                     levels = c('below average', 'average', 'above average'))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
qplot(x = ranking, data = wd)
```

```{r}
summary(wd$ranking)
```

Looking at possibly related variables for trends or relations seems like a 
decent assumption. I'm going to look at sulfur and acidity related values first.
THe relationships between acidity and pH, sugar and alchol content, and alchol
and density as I suspect density in this case is referring to the gravity value
of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = free.sulfur.dioxide, data = wd, binwidth = 1) + xlim(2,40)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$free.sulfur.dioxide)
```

Limiting the x axis lead to removal of 40 records

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = total.sulfur.dioxide, data = wd, binwidth = 1) + 
  scale_x_continuous(limits = c(6, 130), breaks = seq(0, 130, 10))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$total.sulfur.dioxide)
```
Limiting the x axis lead to removal of 41 records

The next few charts relate to the acidity measures.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = fixed.acidity, data = wd, binwidth = .25) + 
  scale_x_continuous(limits = c(4.5, 16), breaks = seq(4.5, 16, 0.5))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$fixed.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = volatile.acidity, data = wd, binwidth = .02) + 
  scale_x_continuous(limits = c(0.1, 1.05), breaks = seq(0.1, 1.1, 0.1))
```
Limiting the x axis lead to removal of 10 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$volatile.acidity)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = citric.acid, data = wd, binwidth = .01) + 
  scale_x_continuous(limits = c(0, 0.75), breaks = seq(0, 0.75, 0.1))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$citric.acid)
```

Limiting the x axis lead to removal of 6 records.

The next set of plots deal with residual sugars, alchol, and density

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = residual.sugar, data = wd, binwidth = .1) + 
  scale_x_continuous(limits = c(0.9, 9), breaks = seq(0.9, 9, 0.5))
```

Limiting the x axis lead to removal of 10 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$residual.sugar)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = alcohol, data = wd, binwidth = .1) + 
  scale_x_continuous(limits = c(8.9, 13.5), breaks = seq(8.9, 13.5, 0.5))
```

Limiting the x axis lead to removal of 20 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$alcohol)
```

This is another measure related to sugar to alchol conversions in
terms of fermentation potential and actual. The vinter would usually take an 
originial gravity reading and a final gravity, what I believe density to be, to
measure fermentable effeciency.


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = density, data = wd, binwidth = .0001) + 
  scale_x_continuous(limits = c(0.99, 1.01), breaks = seq(0.99, 1.003, 0.005))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$density)
```
The last group of variables or other measures that initially I do not know 
or have any expectations about their respective values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = chlorides, data = wd, binwidth = .0001) + 
  scale_x_continuous(limits = c(0.012, 0.162), breaks = seq(0.012, 0.162, 0.02))
```

Limiting the x axis lead to removal of 59 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$chlorides)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = pH, data = wd, binwidth = .01) + 
  scale_x_continuous(limits = c(2.7, 3.7), breaks = seq(2.7, 3.5, 0.1))
```

Limiting the x axis lead to removal of 16 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$pH)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = sulphates, data = wd, binwidth = .01) + 
  scale_x_continuous(limits = c(0.33, 1.2), breaks = seq(0.33, 1.2, 0.1))
```

Limiting the x axis lead to removal of 18 records.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wd$sulphates)
```


#Univariate Analysis

#Template Questions
#-------------------------------------------------

##What is the structure of your dataset?
The Red wine data set has 1599 items and 13 variables for each entry, one of 
which is a simple numerical listing of all the entries.  There is one subjective
value, quaility, which is how the wine was judged based on its othere values. 
The rest of the data set is composed of the chemical measurements of the wine.
The data doesn't contain one variable that may become a lurking variable, taste 
notes. I 

##What is/are the main feature(s) of interest in your dataset?
The features that have my interest relate my an existing hobby and area of 
interest I already have. I'm more interested in the fermentation related data 
including sugar, alchol, and density. My other area of interests is looking at 
the quality scores given and how it relates to the other variables.

##What other features in the dataset do you think will help support your \
##investigation into your feature(s) of interest?
At this point I think there will be common trends in the wines that fall into 
the 3 different ranking groups. Using the chemical properties, I suspect some 
things will be found to be indicitive of an above average wine.

##Did you create any new variables from existing variables in the dataset?
I created a ranking of the quality values into 3 categories of below average, 
average, and above average. I may have to create more once I get into the 
bivariate exploration.

##Of the features you investigated, were there any unusual distributions?
Quality, density, chlorides, and volatile acidity have more uniform 
distributions. Both sulfur measures, sugar, alchol, fixed acidity, and sulfates
have a some sort of leftward skew. pH right skewed slightly right.
Surprisingly citric acid has a flatter but still slightly skewed left 
distribution.

##Did you perform any operations on the data to tidy, adjust, or change the form\
##of the data? If so, why did you do this?
The data appears pretty tidy and easy to work with. I haven't had to change any
forms yet but have to do some adjustments to compare some values more easily.


#Bivariate Plots Section

Next up, I'm going to compare different chemical properties against the 
subjective quality measure and against other chemical markers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = ranking), data = wd, ) + geom_point(alpha = 1/2, color = 'purple')
```

Seems a bit too tidy and ordered, going to jitter the plot to so the density of the data a bit better.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = ranking), data = wd, ) + geom_point(alpha = 1/2, position = 'jitter', color = 'blue') 
```

That is better in terms of relecting the data in each category.

I find it interesting that it appears the above and below average rankings had 
both have a more narrow band in their repective alcohol content. 

I'm going to continue plotting out the various sets I'm curious about.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = quality), data = wd) + geom_point(alpha = 1/2) 
```


```{r echo=FALSE}
ggplot(aes(x = pH, y = ranking),  data = wd) + geom_point(alpha = 1/2, , position = 'jitter') 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = ranking, y = citric.acid), data = wd)  + geom_point(alpha = 1/5, position = 'jitter') 
```

I'm going to add jitter properties to plots that appear too clustered and possibly some trend lines if they make sense for the data points.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol), data = wd) + geom_point(alpha = 1/20)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = density), data = wd) + geom_point(alpha = 1/20)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = density), data = wd) + geom_point(alpha = 1/10, color = 'red')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = chlorides, y = sulphates), data = wd) + geom_point(alpha = 1/20)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide, y = sulphates), data = wd) + geom_point(alpha = 1/10, color = 'orange') + geom_line(stat = 'summary', fun.y = mean) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), data = wd) + geom_point(alpha = 1/20, color = 'green') + geom_line(stat = 'summary', fun.y = mean) 
```


#Bivariate Analysis

The first set of information that I want to check is the correlation values between the campared values. Using a pearson method caluculation as shown below I've calucated all the below values.

I'm going to have to use the raw quality scores that ranking is derived from.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(wd$alcohol, wd$quality, method = 'pearson')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(wd$alcohol, wd$quality, method = 'pearson')

cor.test(wd$pH, wd$quality, method = 'pearson')

cor.test(wd$citric.acid, wd$quality, method = 'pearson')

cor.test(wd$residual.sugar, wd$alcohol, method = 'pearson')

cor.test(wd$residual.sugar, wd$density, method = 'pearson')

cor.test(wd$alcohol, wd$density, method = 'pearson')

cor.test(wd$chlorides, wd$sulphates, method = 'pearson')

cor.test(wd$free.sulfur.dioxide, wd$sulphates, method = 'pearson')
cor.test(wd$free.sulfur.dioxide, wd$total.sulfur.dioxide, method = 'pearson')
```

Below is a table of the Correlation between data variables.



Variable 1 | Variable 2 | r | Strength | -/+
Alcohol | Quality | 0.48 | Moderate | +
pH | Quality | -0.06 | Weak | -
Citric Acid | Quality | 0.23 | Weak | +
Sugar | Alcohol | 0.04 | Weak | +
Sugar | Density | 0.36 | Moderate | +
Alcohol | Density | -0.49 | Moderate | -
Cholrides | Sulphates | 0.37 | Moderate |+
Free Sulfur | Sulphates  | 0.05 | Weak | +
Free Sulfur | Total Sulfur | 0.67 | Strong | +


I am surprised by a few of the correlations. The free sulfur dioxide and total 
sulfur dioxide have a strong positive relationship that isn't surprising as 
its measuring the same chemical compounds. Citric acid and pH don't have a 
relation to quality scores, so I can assume the judges probably had no biases 
to these qualities of the wine. 

Sugar and density and alcohol and density both had moderate r values but the 
measures had opposite polarities. Sugar being the fuel for the yeast
is responsible for both measures and their directions. Density is the relational
measure offermentation potential and the goal range is 0 or below. Alcohol is 
less dense than water and sugar so a negative value denotes higher conversion. 
This is also shown in the moderate to strong negative correlation between 
alcohol and density with a r of -0.49, meaning more alcohol means a lower 
density but sugar leads to a higher value.

It's an interesting note in trying to objectively explain the judges quality 
score, that judges statistically are more inclined to rate a higher ABV with a 
higher score.



### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

My main point of interest are factors that may show trends in the taste judging 
of the samples. It seems most chemical markers have insignificant relations but
there was a significant r in correlating alcohol with the scores.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)? 
From my own interests, I found the mathematical relationship that shows the 
sugar, alcohol, and density interesting to see charted out.

### What was the strongest relationship you found?
The correlation of between sulfur measurements.

# Multivariate Plots Section

To start off the multivariate investigation, it seems sound to some matrices 
plots and look for any additional investigations that haven't been apparent to 
me so far. I've created a subset to exclude the sequential sample numbers of the
wines X.

```{r}
nox <- subset(wd, select=c(               
fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality, ranking))
```

Using this subset the matrix is as follows.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(GGally)
set.seed(1599)
names(nox)
ggpairs(nox)
```


I don't see anything thats super unexpected. The residual sugar and alcohol both skew the same
as expected. The citric acid to sugar distribution is a little wider than I had thought but that sharp decline on the tail is interesting, so I'm going to add that to investigative plots.

From the previous data explorations, the first relationship I want to look at
is the one between residular sugar, alcohol, and quality/ranking.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol), data = wd) +
  geom_line(aes(color = quality), stat = 'summary', fun.y = median)
```

That isn't a view that is easily readable by most, so if I use the hard set 
ranking categories it may translate better to the veiwer.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol), data = wd) +
  geom_line(aes(color = ranking), stat = 'summary', fun.y = median)
```

Ok that makes the rankings stand out a bit more and also shows that above the 
entries judged with a higher score overall plot higher in alcohol. 

I want to explore some of the more posible complex chemical relations. Im just 
trying out a some different combinations looking for things that stand out. The 
citric acid and sugar relation caught my eye in the mass plotting so I want to 
investigate that.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = citric.acid), data = wd) +
  geom_line(aes(color = ranking), stat = 'summary', fun.y = median)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = residual.sugar/100), data = wd) +
  geom_line(aes(color = ranking), stat = 'summary', fun.y = median)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = volatile.acidity), data = wd) +
  geom_bar(aes(color = citric.acid), stat = 'summary', fun.y = mean)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = sulphates), data = wd) +
  geom_bar(aes(color = ranking), stat = 'summary', fun.y = median)
```

Interesting view of two no related chemical markers in respect to quality score.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = chlorides, y = sulphates),
       data = wd) + geom_boxplot() + 
  stat_summary(fun.y = mean, geom = 'point', shape = 4)
```

Well that doesn't really show much. I'm going to keep exploring different measures to quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = total.sulfur.dioxide, y = citric.acid), data = wd) +
  geom_bar(aes(color = ranking), stat = 'summary', fun.y = mean)
```

That's more like the views Im hoping to find. Going to note that the highest 
sulfur samples happened to rank in the top group. I restricted the axis for 
better presentation so those values were lost.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = total.sulfur.dioxide, y = citric.acid), data = wd) +
  geom_bar(aes(color = ranking), stat = 'summary', fun.y = mean) + xlim(0, 160)
```




```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = sulphates), data = wd) +
  geom_bar(aes(color = ranking), stat = 'summary', fun.y = median) + xlim(0, .77)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y = alcohol, x = residual.sugar, fill = quality), data = nox) + geom_tile() + scale_fill_gradientn(colors = colorRampPalette(c('green', 'red'))(100))
```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The scieniftic explanation of alcohol, sugar, and density plot out nicely as 
expected. It appears that there are trends in higher rated wines, may not a 
chemical composistion but just the quanity that lead to a better taste 
experience. Knowing the quality ranking criteria and categorical scores but be
a great data point to have in furthering this investigation.

### Were there any interesting or surprising interactions between features?
The overall measures of most chemical existing in higher rates in the highest 
rated samples has my curiousity peaked in this data. Maybe scores can't be 
factored out into key points, simply the higher content but still in relation to
the other compounds make flavor even more complicated to account for.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE}

set.seed(1599)
names(nox)
ggpairs(nox)

```

### Description One
This is the overall scatterplot matrix for the data set. This can be argued that
it may be very audience specific to the inclusion in presenting data with this 
added to it. For those that understand what is shown it adds an overall insight 
to the relations explored.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = residual.sugar/100), data = wd) +
  geom_line(aes(color = ranking), stat = 'summary', fun.y = median)
```


### Description Two

This plot helps point to a recurring trend that higher quality wines average a 
lower resisudal sugar content. Some assumptions can be made to that line of 
thinking that could open a deeper analysis with a more robust data set. If I was
the vinter of some juddged wines and had additional data such as age, grape 
variety, and other variables of the winery, this could help me improve the 
quaility or drive to better consistent products that are likely to be favored in
judgings.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y = alcohol, x = residual.sugar, fill = quality), data = nox) + geom_tile() + scale_fill_gradientn(colors = colorRampPalette(c('green', 'red'))(100))
```
### Description Three

This a related plot to some of the other discussions showing 'dryer', lower 
sweetness, and higher ABV measures tend to be favored more. 


# Reflection


I found this data set interesting but maybe a bit incomplete for the questions 
it aroused in myself. My largest interest is how this can be used to find the
factors that impact judging of the wines. There were some trends and relations
that were found that helped shed some light on the question at hand. The most 
common factor seem the concentration of compound including alcohol tend to score
higher. This could be due to several factors that prompt further discovery. I
feel that having the values that were judged to get the quality scores, as well
vinter specific inforamtion such as region, yeast strain, grape variety, and age
could really round out the picture of what goes into making great wine.